Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-05-24 manuscript/manuscript.tex
Modifications in Determinations of D and delta

The discretization algorithms described on the previous section are 
introduced to improve the accuracy of the numerical approximation of 
the gravitational fields.
They involve two predefined ratios that determine how many 
times each tesseroid will be discretized and thus indirectly control 
the numerical error of the approximation.
The modified version of the adaptive discretization algorithm involves  
the distance-size ratio $D$, while the density-based discretization 
algorithm makes use of the delta ratio $\delta$.
These algorithms need default predefined values for $D$ and $\delta$ 
in order to ensure both acceptable numerical accuracy and computation 
effectiveness.

@@

\noindent where $\ell$ is defined in equation \ref{eq:ell}.
The computation point $Q$ is located at latitude $\phi=90^\circ$, so 
$\ell$ and $\cos\psi$ (defined in equation \ref{eq:cospsi}) get 
simplified:

@@

Firstly we need to solve the integral on equation \ref{eq:shell-pot} in 
order to obtain the analytical solution in this case.
Thus, the gravitational potential generated by a spherical shell with this 
density on any external point at distance $r$ from its centre can be 
obtained from the following expression:
\todo[inline]{revisar redaccion}

@@

\noindent The first term on equation \ref{eq:shell-pot-linear} 
reproduces the potential generated by a spherical shell with variable 
density $\rho(r') = ar'$, while the second one constitutes the 
potential generated by a spherical shell with homogeneous density $\rho 
= b$ \citep{Mikuska2006,Grombein2013}.

@@

In order to compare the numerical model with the analytical solution we 
must build a spherical shell from a set of tesseroids.
We propose a mesh of $30^\circ \times 30^\circ$ tesseroids to compute 
the gravitational potential, the vertical component of the gradient 
($g_z$) and the diagonal components of the Marussi tensor ($g_{xx}$, 
$g_{yy}$, $g_{zz}$) on four different grids: (1) a grid located at the 
pole, (2) another one on the Equator, (3) one located at satellite 
height and (4) a big grid of $30^\circ \times 30^\circ$.
More information about this grids can be found on Table 
\ref{tab:grids}.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model 
    and the analytical solution for a thin and a thick spherical shell of 1km 
    and 35km of thickness, respectively.
    Both models have the linear density defined on equation 
    \ref{eq:density-linear}.
    The computations were performed on the four grids described in 
    Table \ref{tab:grids}, for different values of the distance-size 
    ratio $D$.
    Due to the linearity of the density function the density-based 
    discretization algorithm is not applied.
    If the difference is less than 0.1\%, we consider that the model 
    has achieved an acceptable accuracy for the corresponding value of 
    $D$.
    }
\label{fig:D-linear}
\end{figure*}

@@

@@

\noindent $R$ is the mean Earth radius and $b$ is a constant 
that determines the variability of the function: a low value of $b$ 
increases the maximum slope of the density function.

The analytical solution of the gravitational fields generated 
by a spherical shell with a density functions as the one in equation 
\ref{eq:density-exp} can be computed as the sum of the fields generated 
by a spherical shell with constant density $\rho(r') = C$ and another 
shell with an exponential density like:

\begin{equation}
    \rho(r') = Ae^{-(r' - \Delta h)/b},
\end{equation}

\noindent where $A$, $\Delta h$ and $b$ are constants.
The gravitational potential field generated by a spherical shell with 
the latter density on any external point at a distance $r$ from its centre 
can be calculated by solving the integral in equation \ref{eq:shell-pot},
resulting as follows:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
        Ab 
        \Big[
        & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1 - \Delta h}{b}} - \\
        & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2 - \Delta h}{b}}
        \Big].
    \end{split}
\label{eq:shell-pot-exp}
\end{equation}
}{
\begin{equation}
    V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
    Ab \, e^\frac{\Delta h}{b}
    \Big[
    (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} - \\
    (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
    \Big].
\label{eq:shell-pot-exp}
\end{equation}
}

@@

On Figure \ref{fig:grid-search} we show 
the resulting percentage difference for each gravitational field.
The points inside the dotted lines are the ones that present a 
numerical error lower than the 0.1\% threshold.
The one that produces the more efficient computation must be chosen as 
the default $D$ and $\delta$ value.

@@

Because the algorithm is intended to work with default $D$ and $\delta$ 
values that ensures these two conditions for every density function, 
the default value for $D$ cannot be lower than the one determined for 
the linear function, which will be called $D_\text{linear}$.
That's why we can only choose the ($D$, $\delta$) points on which
$D \geq D_\text{linear}$.
Taking this fact into account we can divide the $D$, $\delta$ space in 
two subregions: one with $D < D_\text{linear}$ and other with $D \ge 
D_\text{linear}$. We can ignore points belonging to the former one 
(region A on Figure \ref{fig:D-delta-grid-search}) and focus on the 
latter (regions B and C on Figure \ref{fig:D-delta-grid-search}).

@@

>>>>

The discretization algorithms described on the previous section are 
introduced to improve the accuracy of the numerical approximation of 
the gravitational fields.
They involve two predefined ratios that determine how many 
times each tesseroid will be discretized and thus indirectly control 
the numerical error of the approximation.
The modified version of the adaptive discretization algorithm involves  
the distance-size ratio $D$, while the density-based discretization 
algorithm makes use of the delta ratio $\delta$.
These algorithms need predefined default values for $D$ and $\delta$ 
in order to ensure both acceptable numerical accuracy and computation 
effectiveness.

@@

\noindent where $\ell$ is defined in equation \ref{eq:ell}. The computation 
point $Q$ is located at latitude $\phi=90^\circ$, so $\ell$ and $\cos\psi$ 
(defined on equations \ref{eq:ell} and \ref{eq:cospsi}) get simplified:

@@

First, by solving the integral on equation \ref{eq:shell-pot}, we obtained the 
gravitational potential generated by a spherical shell with this density on 
any external point at distance $r$ from its centre:

@@

\noindent The first term on this equation reproduces the potential generated 
by a spherical shell with variable density $\rho(r') = ar'$, while the second 
one constitutes the potential generated by a spherical shell with homogeneous 
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.

@@

In order to compare the numerical model with the analytical solution we 
must build a spherical shell made of tesseroids.
We propose a mesh of $30^\circ \times 30^\circ$ tesseroids whose outer and 
inner radii are the same ones as the spherical shell in order to compute 
the gravitational potential, the vertical component of the gradient 
($g_z$) and the diagonal components of the Marussi tensor ($g_{xx}$, 
$g_{yy}$, $g_{zz}$) on four different grids: (1) a grid located at the 
pole, (2) another one on the Equator, (3) one located at satellite 
height and (4) a big grid of $30^\circ \times 30^\circ$.
More information about this grids can be found on Table 
\ref{tab:grids}.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model 
    and the analytical solution for a thin and a thick spherical shell of 1km 
    and 35km of thickness, respectively.
    Both models have the linear density defined on equation 
    \ref{eq:density-linear}.
    The computations were performed on the four grids described in 
    Table \ref{tab:grids}, for different values of the distance-size 
    ratio $D$.
    Due to the linearity of the density function, the density-based 
    discretization algorithm is not applied.
    If the difference is less than 0.1\%, we consider that the model 
    has achieved an acceptable accuracy for the corresponding value of 
    $D$.
    }
\label{fig:D-linear}
\end{figure*}

@@

\subsubsection{$D$, $\delta$ space exploration}

@@

\noindent $R$ is the mean Earth radius and $b$ is a constant 
that determines the variability of the function: a low value of $b$ 
increases the maximum slope of the density function.
The analytical solution of the gravitational potential generated by a 
spherical shell with a density functions as the one in equation 
\ref{eq:density-exp} can be obtained by solving the equation 
\ref{eq:shell-pot} as follows:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
        Ab 
        \Big[
        & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1 - \Delta h}{b}} - \\
        & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2 - \Delta h}{b}}
        \Big] + \\
        & \frac{4 \pi G}{3 r} C (R_2^3 - R_1^3).
    \end{split}
\label{eq:shell-pot-exp}
\end{equation}
}{
\begin{equation}
    V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
    Ab \, e^\frac{\Delta h}{b}
    \Big[
    (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} -
    (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
    \Big] +
    \frac{4 \pi G}{3 r} C (R_2^3 - R_1^3).
\label{eq:shell-pot-exp}
\end{equation}
}

@@

On Figure \ref{fig:grid-search} we show 
the resulting percentage difference for each gravitational field.
The points inside the dotted lines are the ones that present a 
numerical error lower than the 0.1\% threshold.
Our intention is to choose the one that produces the most efficient computation
and set it as the default values for $D$ and $\delta$.

@@

Because the algorithm is thought to work with default $D$ and $\delta$ 
values that ensures these two conditions for every density function, 
the default value for $D$ cannot be lower than the one determined for 
the linear function, which will be called $D_\text{linear}$.
That's why we can only choose the ($D$, $\delta$) points on which
$D \geq D_\text{linear}$.
Taking this fact into account we can divide the $D$, $\delta$ space in 
two subregions: one with $D < D_\text{linear}$ and other with $D \ge 
D_\text{linear}$. We can ignore points belonging to the former one 
(region A on Figure \ref{fig:D-delta-grid-search}) and focus on the 
latter (regions B and C on Figure \ref{fig:D-delta-grid-search}).

@@

\subsubsection{Delta ratio determination}
